---
title: "Reproducible Research: Peer Assessment 1"
Author: Johnny Sandaire
output: 
  html_document:
    keep_md: true
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(scales)
library(Hmisc)
library(dplyr)
library(lubridate)
```

## Loading and preprocessing the data

##### 1. Load the data (i.e. read.csv())

```{r, results='markup'}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
walkData <- read.csv('activity.csv')
summary(walkData)
```

## What is mean total number of steps taken per day?
```{r}
dailySteps <- walkData %>% 
      group_by(date) %>%
      summarise(totalSteps=sum(steps, na.rm = TRUE))
```

##### Make a histogram of the total number of steps taken each day
```{r}
hist(x=dailySteps$totalSteps, main = "Total number of steps taken per day", 
     xlab = "Total steps taken per day", ylab = "Number of days", 
     breaks = 10, col = "red")
abline(v = mean(dailySteps$totalSteps, na.rm = TRUE), lty = 1, lwd = 2, col = "steel blue")
abline(v = median(dailySteps$totalSteps, na.rm = TRUE), lty = 2, lwd = 2, col = "yellow")
text(y = 15, x = mean(dailySteps$totalSteps, na.rm = TRUE), pos=2, labels = "Mean")
text(y = 15, x = median(dailySteps$totalSteps, na.rm = TRUE), pos=4, labels = "Median", col="black")
```

##### Calculate and report the mean and median total number of steps taken per day
```{r}
dailyStepsMean <- mean(dailySteps$totalSteps, na.rm = TRUE)
dailyStepsMedian <- median(dailySteps$totalSteps, na.rm = TRUE)
```
1.  Mean Total Number of Steps Taken Per Day: `r dailyStepsMean`
2.  Median Total Number of Steps Taken Per Day:  `r dailyStepsMedian`

-----

## What is the average daily activity pattern?
```{r}
avgDailyWalkPat <- aggregate(x=list(meanWalk=walkData$steps), by=list(interval=walkData$interval), FUN=mean, na.rm=TRUE)
```

##### Make a time series plot
```{r}
ggplot(data=avgDailyWalkPat, aes(x=interval, y=meanWalk)) +
    geom_line() +
    xlab("5-minute Walk Interval") +
    ylab("Mean Number of Steps") 
```

##### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
mostSteps <- which.max(avgDailyWalkPat$meanWalk)
timeMostSteps <-  gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", avgDailyWalkPat[mostSteps,'interval'])
```

* Most Steps at: `r timeMostSteps`

----

## Imputing missing values
##### Calculate and report the total number of missing values in the dataset 
```{r}
numMissingValues <- length(which(is.na(walkData$steps)))
```

* Number of missing values: `r numMissingValues`

##### Devise a strategy for filling in all of the missing values in the dataset.
##### Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
walkDataImputed <- walkData
walkDataImputed$steps <- impute(walkData$steps, fun=mean)
```


##### Make a histogram of the total number of steps taken each day 
```{r}
dailyStepsImputed <- walkDataImputed %>% 
      group_by(date) %>%
      summarise(total_steps=sum(steps, na.rm = TRUE))
hist(x=dailyStepsImputed$total_steps, main = "Total number of steps taken per day", 
     xlab = "Total steps taken per day", ylab = "Number of days", 
     breaks = 10, col = "red")
abline(v = mean(dailyStepsImputed$total_steps, na.rm = TRUE), lty = 1, lwd = 2, col = "yellow")
abline(v = median(dailyStepsImputed$total_steps, na.rm = TRUE), lty = 2, lwd = 2, col = "black")
text(y = 15, x = mean(dailyStepsImputed$total_steps, na.rm = TRUE), pos=4, labels = "mean")
text(y = 15, x = median(dailyStepsImputed$total_steps, na.rm = TRUE), pos=2, labels = "median")
```


##### Calculate and report the mean and median total number of steps taken per day. 
```{r}
dailyStepsMeanImputed <- mean(dailyStepsImputed$total_steps, na.rm = TRUE)
dailyStepsMedianImputed <- median(dailyStepsImputed$total_steps, na.rm = TRUE)

```
* Mean (Imputed): `r dailyStepsMeanImputed`
* Median (Imputed):  `r dailyStepsMedianImputed`


----

## Are there differences in activity patterns between weekdays and weekends?
##### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
walkDataImputed$dateType <-  ifelse(as.POSIXlt(walkDataImputed$date)$wday %in% c(0,6), 'weekend', 'weekday')
```

##### Make a panel plot containing a time series plot

```{r}
averagedwalkDataImputed <- aggregate(steps ~ interval + dateType, data=walkDataImputed, mean)
ggplot(averagedwalkDataImputed, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(dateType ~ .) +
    xlab("5-minute Walk Interval") + 
    ylab("Mean Number of Steps")
```
